---
title: "Final_Honors_Thesis"
author: "Jeannie Hinton"
date: "2024-10-27"
output: html_document
---
#Loading Libraries
```{r loading libraries}
knitr::opts_chunk$set(echo = TRUE)
#loading necessary libraries
library(tidyverse)
library(fs)
library(corrr)
library(tidymodels)
```


#Wrangling the Data for Processing
```{r setting up data for wrangling, eval=FALSE}
#making a file path to be stepped through for a data list below
file_paths <- dir_ls("raw_data/train_tsv", glob = "*.tsv")

#making a list that stores all of the correlation matrices
#in an effort to be able to index and extract the matrices for processing
data_list_train <- file_paths %>%
  map(~ read_tsv(.x, col_names = FALSE))

#saving data_list as an object for use later
saveRDS(data_list_train, file = "data/data_list_train.rds")
```

##Background Work
```{r background work, eval = FALSE}
#loads the data_list_train object from the data file for processing
data_list_train <- readRDS(file = "data/data_list_train.rds")
#view(data_list_train[[1]])
#view(data_list_train[[1104]])
```

```{r create vector of file names, eval = FALSE}
file_names_list <- list.files(path = "raw_data/train_tsv", pattern = ".tsv", all.files = TRUE, full.names = TRUE)
simplified_names_list <- sapply(str_split(file_names_list, "_"),'[', 3)
participant_id_list <- gsub('tsv/sub-', '', simplified_names_list)
```

```{r extracting correlations and loading into vectors, eval = FALSE}
# Creating an empty list for storing correlation values
corr_vals_id_list <- list()

# Loop to read in correlation values and assign them to variables with indexed names
for (i in seq_along(data_list_train)) {
  # The pluck function extracts the i-th element from the training data_list
  corr1 <- pluck(data_list_train, i)
  
  # Extract the upper triangular elements. This fills the vector elements by column.
  # The function upper.tri creates a vector of true/false values. We then select all the true values.
  upper_vec <- corr1[upper.tri(corr1, diag = FALSE)]
  
  
  # Create a named list combining the participant ids and upper_vec values
  upper_vec_with_id <- list(participant_id = participant_id_list[i], corr_values = upper_vec)
  

  # Assign each id and upper_vec values pair to a position in the list
  corr_vals_id_list[[i]] <- upper_vec_with_id
}

# Check the list worked correctly
#print(corr_vals_id_list[[100]])
#view(data_list_train[[100]])

#store this list of corr values and ids as an r object
saveRDS(corr_vals_id_list, file = "data/corr_vals_id_list.rds")
```

```{r changing list to tibble, eval = FALSE}
# first list element is ID, second is vector of correlations
# need to put in a dataframe
list_all <- read_rds("data/corr_vals_id_list.rds")

# created vector with column names for the correlation values:
names <- paste0("X", seq(1:19900))
names <- c("id", names)
 
# pass this vector length to ncol parameter
# and nrow with 0
corr_data = data.frame(matrix(nrow = 0, ncol = length(names)))
 
# assign column names
colnames(corr_data) = names

# loop through each participant
for (i in 1:1104) {
#grab the next element in the list
 listrow <- pluck(list_all, i)
 corr_data[i, 1] <- listrow$participant_id
 corr_data[i, 2:19901] <- listrow$corr_values #assigning the vector elements to each col
}

#getting error =  Error in x[[jj]][iseq] <- vjj : replacement has length zero


#check first and last ids
##corr_data[[2]]
```


```{r loads r object of corr values and id df}
#loads the corr_vals_id_df object from the data file for processing
corr_data <- readRDS(file = "data/corr_data.rds")
```

```{r checking corr_data quality}
sum(is.na(corr_data[[1]]))
sum(is.null(corr_data[[1]]))

#EDA ON TRAINING DATA
#using cross validation for tuning and also to get a rmse estimate for each model that i fit
#missing values in recipe
#zero variance 
#dummy variables 
#start the recipe
#going down correlation matrix
#join table and start doing basic visualizations, 5 number summary of a few correlation column
#pca research, research regularized regression or pca before fitting 
#caret train or do a workflow
```


##MetaData Work

```{r}
train_meta <- read_csv("raw_data/metadata/training_metadata.csv")
train_meta <- train_meta %>% 
  rename(id = participant_id)
#view(train_meta)
#view(train_meta)
sum(is.na(train_meta))

train_meta %>% 
  skimr::skim(.)
#not enough NAs to drop any of these columns entirely

train_meta %>% 
  count(handedness)
```

```{r merging tables}
#joining meta data to the correlation data to create one large dataframe 
full_table <- left_join(train_meta, corr_data, by = "id")

#changing categorical predictors from char to factors
full_table <- full_table %>% 
  mutate(id = as.factor(id),
         sex = as.factor(sex),
         study_site = as.factor(study_site),
         ethnicity = as.factor(ethnicity),
         race = as.factor(race),
         handedness = as.factor(handedness), 
         parent_1_education = as.factor(parent_1_education), 
         parent_2_education = as.factor(parent_2_education))
  
#verifying that the changing of data types worked correctly
# #full_table %>% 
#   select(1:14) %>% 
#   glimpse()
  
#saving data_list as an object for use later
saveRDS(full_table, file = "data/full_table.rds")

#checking to make sure missing values align with what was missing in meta data and not missing in 
#the correlation data
#sum(is.na(full_training_table))
```

```{r load in final table}
full_table <- readRDS(full_table, file = "data/full_table.rds")
```

```{r split final table}
set.seed(123)

#put 80% of the training data into the training set 
full_data_split <- initial_split(full_table, prop = 0.80)

#create data frames for the two sets:
train_split <- training(full_data_split)
test_split  <- testing(full_data_split)

#create r objects to store the training and test splits
#this is really in an effort to seperate test data entirely into a different file to prevent any 
#mistaken manipulation of it while manipulating the training data
saveRDS(train_split, file = "data/train_split.rds")
saveRDS(test_split, file = "test_data_dont_touch/test_split.rds")

```



```{r reading in training split}
train_split <- readRDS(train_split, file = "data/train_split.rds")
```



#Preprocessing the Data 
```{r preprocessing data}

#do we know how the metadata was collected? survey? impacts how to deal with missing values

#confirming that changed data types carried through into the split of the data correctly
#train_split %>% 
  # select(1:15) %>% 
  # glimpse()


#handle missing values
corr_data_rec <- recipe(age ~ ., data = train_split) %>% 
  step_rm(id) %>% 
  step_impute_mean() %>% 
  step_impute_mode()

broom::tidy(corr_data_rec)
#note that I will not be removing any of these predictors, because the typical threshold for throwing out a column that has missing values is 
#20% of a predictors' observations should be missing in order to throw out the predictor. This is not the case. Make note of this in write up


##you can start by trying a simple approach:  replace with the mean for quantitative variables; replace ##with the mode for categorical variables.
```
| Predictor | # Missing Values
|:-------|:-----------
| bmi  | 18             
| ethnicity   | 109             
| race  | 168            
| parent_1_education | 183            
| parent_2_eduction  | 22               


```{r models}
# corr_data_wflow <- 
#   workflow() %>% 
#   add_model() %>%  #need to add a model in here
#   add_recipe(corr_data_rec)
```


```{r training the model}
# corr_data_fit <- 
#   corr_data_wflow %>% 
#   fit(data = train_split) %>% 
#   tidy()
```


###General tidymodel steps
Create a recipe for feature engineering steps to be applied to the training data (these steps will also automatically be applied later to the test data)

Fit the model to the training data after applying these steps

Evaluate model/tune model parameters using resampling

Using the model estimated from the training data, predict outcomes for the test data and evaluate performance of the model on the test data



### Feature Engineering

While your project’s needs may vary, here is a suggested order of potential steps that should work for most problems:

1.  Filter out zero or near-zero variance features.

2.  Perform imputation if required.

3.  Normalize to resolve numeric feature skewness.

4.  Standardize (center and scale) numeric features.

5.  Perform dimension reduction (e.g., PCA) on numeric features.

6.  One-hot or dummy encode categorical features.

